Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
fixed bad spacing

\subsection*{Related Work}
Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}. The innovations of Iris (custom ghost state and invariants) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach. Some of the more complex data structures verified include the \lstinline{ConcurrentSkipListMap} from \lstinline{java.util.concurrent}~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}. Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeted at an implementation language (C) rather than a toy language or algorithm-level representation. %do we want to say this part?

@@

Separating concurrency reasoning from data structure reasoning has long been an appealing target. Linearizability~\cite{linearizability}, the most common correctness condition for concurrent data structures, was a first step in this direction, describing the conditions under which a concurrent implementation can replace a sequential one in all possible contexts. Logical atomicity can be seen as a compositional analogue to linearizability, where the proof that each operation implements its sequential counterpart can be carried out independently. Concurrent templates can be seen as the next step of compositionality, allowing the concurrent-to-sequential proof to be decoupled from the specific data structure being verified.

@@

A search structure is a data structure designed to efficiently store and retrieve data based on specific search criteria. In the abstract, a search structure implements a map from keys to values, and provides operations such as search, insertion, deletion, and traversal on that map. For efficiency, many such structures are designed to allow concurrent access and modification, often employing fine-grained or lock-free concurrency to allow as many threads as possible to operate on separate parts of the data structure. Designing these search structures presents significant challenges, including ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory, including cache behavior~\cite{masstree}. The complex interplay between concurrency and data structure design in concurrent search structures makes them challenging targets for formal verification.

@@

Figure~\ref{template-ex} shows an example search structure template. The core of the template is the \lstinline{traverse} function, which uses a particular concurrency control mechanism to travel through a data structure in search of a particular key. The mechanism in this example is \emph{lock coupling}, where we acquire the lock on the next node before releasing the lock on the current node. The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine when the current node does not contain the target key. Once the appropriate node for the key has been found, the template returns it to a top-level function such as \lstinline{insert} that calls out to the data structure to perform the actual insertion on the node. Thus, the \lstinline{traverse} and \lstinline{insert} functions can be verified without knowing anything about the target data structure other than its synchronization mechanism, as long as the data structure implements \lstinline{findNext} and \lstinline{insertOp} operations with the required semantics. The concurrency functions in the template and the sequential \lstinline{findNext} and \lstinline{insertOp} functions provided by the data structure combine into a fully operational concurrent data structure. Krishna et al.~specify templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures (i.e., those in which a node may be reachable along multiple paths), but many common search structures do not require flow-style reasoning; in this paper, we present a version of templates that is independent of flow interfaces. %is this a contribution in itself?
%what more do we want to say here?

@@

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as \\$\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$. Its concurrent counterpart could be specified as 
$$\forall s.\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ (v::s)\right\rangle$$
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

\subsection{Iris and VST}
The search structure template approach uses concurrent separation logic to specify and prove pre- and postconditions for the template and data structure functions. Krishna et al.~implemented their framework in Iris~\cite{iris}, a language-independent CSL framework built in the Coq proof assistant~\cite{coq}, with flexible support for ghost state, invariants, and atomic specifications. This makes it easy to describe the effects of concurrency control functions independently of the underlying data structure, e.g., as ``this function atomically finds a node that contains the target key.'' %forward reference to where we show the specs? or show them here?
The existing implementation is written in Iris's HeapLang, a simple functional programming language with shared-memory concurrency.

>>>>

\subsection*{Related Work}
Recent years have seen major advances in concurrent separation logics (CSLs) for verifying fine-grained concurrent programs, including Iris~\cite{iris}, VST~\cite{plcc,iris-vst-arxiv}, FCSL~\cite{fcsl}, TaDA~\cite{tada}, and VeriFast~\cite{verifast-conc}. The innovations of Iris (custom ghost state) and TaDA (logical atomicity) were particularly essential for the formalization of the search structure template approach. Some of the more complex data structures verified include the \lstinline{ConcurrentSkipListMap} from \lstinline{java.util.concurrent}~\cite{Xiong2017Abstract} and a multi-producer multi-consumer concurrent queue from the Folly library~\cite{iris-folly}. Notably, VST is the only one of these systems that is all three of 1) mechanized (i.e., implemented in a theorem prover), 2) foundational (i.e., connected to a formal semantics of the target language), and 3) targeted at an implementation language (C) rather than a toy language or algorithm-level representation. %do we want to say this part?

@@

Separating concurrency reasoning from data structure reasoning has long been an appealing target. Linearizability~\cite{linearizability}, the most common correctness condition for concurrent data structures, was a first step in this direction, describing the conditions under which a concurrent implementation can replace a sequential one in all possible contexts. Logical atomicity has been shown to be a compositional analogue to linearizability~\cite{la-lin}, where the proof that each operation implements its sequential counterpart can be carried out independently. Concurrency templates can be seen as the next step towards compositionality, allowing us to prove linearizability/atomicity of concurrency patterns independently of specific data structures.

@@

A search structure is a data structure designed to efficiently store and retrieve data based on specific search criteria. In the abstract, a search structure implements a map from keys to values, providing operations such as search, insertion, deletion, and traversal on that map. For efficiency, many search structures allow concurrent access and modification, often employing fine-grained or lock-free concurrency to allow as many threads as possible to operate on separate parts of the data structure. Designing these search structures presents significant challenges, including ensuring correctness and consistency under concurrent access, achieving scalability by minimizing contention and maximizing parallelism, and maintaining performance and efficiency while managing synchronization and memory, including cache behavior~\cite{masstree}. The complex interplay between concurrency and data structure design in concurrent search structures makes them challenging targets for formal verification.

@@

Figure~\ref{template-ex} shows an example search structure template. The core of the template is the \lstinline{traverse} function, which uses a specific concurrency control mechanism to travel through a data structure in search of the requested key. The mechanism in this example is \emph{lock coupling}, where we acquire the lock on the next node before releasing the lock on the current node. The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine. Once the appropriate node for the key has been found, the template returns it to a top-level function such as \lstinline{insert} that calls out to the data structure to perform the actual insertion on the node. Thus, the \lstinline{traverse} and \lstinline{insert} functions can be verified without knowing anything about the target data structure other than its synchronization mechanism, as long as the data structure implements \lstinline{findNext} and \lstinline{insertOp} operations with the required semantics. The concurrency functions in the template and the sequential \lstinline{findNext} and \lstinline{insertOp} functions provided by the data structure combine into a fully operational concurrent data structure. Krishna et al.~specify templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures (i.e., those in which a node may be reachable along multiple paths), but many common search structures do not require flow-style reasoning; in this paper, we present a version of templates that is independent of flow interfaces. %is this a contribution in itself?
%what more do we want to say here?

@@

As an example, a sequential stack \lstinline{push} operation could conventionally be specified as \\$\{\mathsf{stack}\ s\ p\}\ \texttt{push}(v)\ \{\mathsf{stack}\ (v :: s)\ p\}$. Its concurrent counterpart could be specified as 
$$\forall s.\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \mathsf{is\_stack}_g\ p\ |\ \mathsf{stack}_g\ (v::s)\right\rangle$$
indicating that the \lstinline{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transforming the stack from $s$ to $v::s$ at some point during its execution. The stack itself is a shared resource, and can only be accessed and modified atomically by threads holding the corresponding \texttt{is\_stack} assertion. The local and shared assertions are connected by an arbitrary identifier $g$, which we will generally omit when clear from context. Logical atomicity can be used to prove linearizability---if all of a data structure's operations satisfy logically atomic triples derived from the corresponding sequential operations, the data structure is linearizable~\cite{la-lin}---but can also be used to specify more complex behavior of nonlinearizable data structures~\cite{compass}.

@@

\subsection{Iris and VST}
The search structure template approach uses concurrent separation logic to specify and prove pre- and postconditions for the template and data structure functions. Krishna et al.~implemented their framework in Iris~\cite{iris}, a language-independent CSL framework built in the Coq proof assistant~\cite{coq}, with flexible support for ghost state, invariants, and atomic specifications. This makes it easy to describe the effects of concurrency control functions independently of the underlying data structure, e.g., as ``this function atomically finds a node that contains the target key.'' %forward reference to where we show the specs? or show them here?
Algorithms are verified in Iris's HeapLang, a simple functional programming language with shared-memory concurrency.
